---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# `randhclust` project

<!-- badges: start -->
<!-- badges: end -->

This repository contains the `randhclust` R package, which (once this repo is public) can be installed as follows:

```{r}
remotes::install_github("jacobbien/randhclust-project", subdir = "randhclust")
```

It is implemented using literate programming with [litr](https://jacobbien.github.io/litr-project/).  Thus, the source code is in the form of a bookdown, available in [create-randhclust/](create-randhclust/).  For now, to read this as a bookdown, please download the repo and then open [docs/create-randhclust/index.html](docs/create-randhclust/index.html) in a browser.

To modify the code in this R package, modify the `.Rmd` files in the `create-randhclust/` directory and then (from an R session in this directory) run the following:

```{r}
litr::render("create-randhclust/index.Rmd")
fs::dir_copy("create-randhclust/_book", "docs/create", overwrite = TRUE)
fs::dir_delete("create-randhclust/_book")
fs::dir_copy("create-randhclust/randhclust", "randhclust", overwrite = TRUE)
fs::dir_delete("create-randhclust/randhclust")
fs::dir_delete("create-randhclust/_main_files/")
```

To install the latest version of `litr`, run

```{r}
remotes::install_github("jacobbien/litr-project", subdir = "litr")
```

