# Preliminaries

Before getting to the main methodology, let's define some functions for generating example data and making plots.

## Generating data

For $\ell=1,\ldots, k$ and each $i\in C_\ell$, we generate

$$
X_i|Y_i=\ell \sim N_p(\mu_\ell, \sigma^2 I_p).
$$

```{r}
#' Generate samples from a clustering model
#' 
#' Samples X|Y=l ~ N_p(mu_l, sigma^2 * I_p)
#' 
#' @param nums k-vector giving the number of points in each cluster
#' @param mu k-by-p matrix giving the k cluster centers
#' @param sigma within cluster sd
generate_data <- function(nums, mu, sigma = 1) {
  k <- nrow(mu)
  p <- ncol(mu)
  stopifnot(length(nums) == k, nums > 0, sigma > 0)
  x <- matrix(nrow = 0, ncol = p)
  for (l in seq(k)) {
    xl <- matrix(mu[l, ] + sigma * stats::rnorm(nums[l] * p), ncol = p, byrow = TRUE)
    x <- rbind(x, xl)
  }
  list(x = x, clust = rep(1:k, times = nums))
}
```


```{r}
usethis::use_package("stats")
```



Let's generate a simple two-cluster, two-dimensional example, which we'll use throughout.

```{r}
set.seed(123)
ex1 <- generate_data(
  nums = c(40, 20),
  mu = matrix(c(0, 0,
                1, 4), byrow = TRUE, nrow = 2)
  )
```

## Plotting data

```{r}
#' Plot two-dimensional data while showing up to two clusterings
#' 
#' For example, can be used to plot both the true and estimated clusterings.
#' 
#' @param x n-by-2 data matrix
#' @param clust1 n-vector of cluster indices
#' @param clust1_label name of clustering 1 (e.g. "true clustering")
#' @param clust2 n-vector of cluster indices (optional)
#' @param clust2_label name of clustering 2 (e.g. "estimated clustering"); optional
plot_clustered_data <- function(x, clust1, clust1_label = "clust1", clust2 = NULL, clust2_label = "clust2") {
  no_legend <- is.null(clust1_label)
  if (no_legend) clust1_label <- "clust1"
  dat <- tibble::tibble(x1 = x[, 1], x2 = x[, 2], clust1 = clust1)
  dat <- dat %>% 
    dplyr::mutate(!!clust1_label := paste0("clust", clust1))
  if (is.null(clust2)) {
    # show only one clustering
    p <- dat %>% 
      ggplot2::ggplot(ggplot2::aes(x = .data$x1,
                                   y = .data$x2,
                                   color = .data[[clust1_label]])) + 
      ggplot2::geom_point() +
      ggplot2::coord_equal()
  }
  else {
    # show two clusterings
    dat <- dat %>% 
      dplyr::mutate(!!clust2_label := paste0("clust", clust2))
    p <- dat %>% 
      ggplot2::ggplot(ggplot2::aes(x = .data$x1, y = .data$x2,
                 color = .data[[clust1_label]],
                 shape = .data[[clust2_label]])) + 
      ggplot2::geom_point() +
      ggplot2::coord_equal()
  }
  if (no_legend) p <- p + ggplot2::theme(legend.position = "none")
  return(p)
}
```


```{r}
usethis::use_pipe()
usethis::use_import_from("rlang", ".data")
usethis::use_package("tibble")
usethis::use_package("dplyr")
usethis::use_package("ggplot2")
```

Let's try this when we give it only the true clustering:

```{r}
plot_clustered_data(ex1$x, ex1$clust, clust1_label = "true clustering")
```
Now let's perform hierarchical clustering and compare the true to estimated:

```{r}
ex1$d <- dist(ex1$x)
ex1$hc_complete$hc = hclust(ex1$d, method = "complete")
plot_clustered_data(
  ex1$x,
  clust1 = ex1$clust,
  clust1_label = "true clustering",
  clust2 = cutree(ex1$hc_complete$hc, k = 2),
  clust2_label = "estimated clustering")
```


