# Randomized Hierarchical Clustering

## Define some linkages

Given a dissimilarity matrix $d\in\mathbb R^{n\times n}$, a linkage is a function that takes two index sets $C_1$ and $C_2$ and outputs a non-negative number (representing the dissimilarity of the two clusters).

```{r linkage-params}
#' @param clust1 a vector giving a subset of `1:n`
#' @param clust2 a vector giving a subset of `1:n`
#' @param d an n-by-n matrix of dissimilarities
```


### Complete linkage

$$
d(C_1,C_2)=\max_{i\in C_1,j\in C_2}d_{ij}
$$

```{r}
#' Complete linkage
#' 
<<linkage-params>>
linkage_complete <- function(clust1, clust2, d) {
  max(d[clust1, clust2])
}
```


### Single linkage

$$
d(C_1,C_2)=\min_{i\in C_1,j\in C_2}d_{ij}
$$

```{r}
#' Single linkage
#' 
<<linkage-params>>
linkage_single <- function(clust1, clust2, d) {
  min(d[clust1, clust2])
}
```

### Group-average linkage

$$
d(C_1,C_2)=\frac1{|C_1|\cdot|C_2|}\sum_{i\in C_1,j\in C_2}d_{ij}
$$
```{r}
#' Group average linkage
#' 
<<linkage-params>>
linkage_group_average <- function(clust1, clust2, d) {
  mean(d[clust1, clust2])
}
```

### Minimax linkage

$$
d(C_1,C_2)=\min_{i\in C_1\cup C_2}\max_{j\in C_1\cup C_2}d_{ij}
$$
```{r}
#' Minimax linkage
#' 
<<linkage-params>>
linkage_minimax <- function(clust1, clust2, d) {
  both <- c(clust1, clust2)
  min(apply(d[both, both], 1, max))
}
```

## Randomized version of `hclust` {#randomized-hclust}

We begin by writing a function for computing the merge probabilities as in Section 2 of the paper:

$$
p_{kl} =\begin{cases}
\frac{\exp(-d(C_k,C_l)/\tau)}{\sum_{k'<l'}\exp(-d(C_{k'},C_{l'})/\tau)}& \text{if } \tau>0\\
|A|^{-1}\cdot1\{(C_k,C_l)\in A\} & \text{if }\tau=0
\end{cases}
$$
where $A=\arg\min_{k'l'}d(C_{k'},C_{l'})$.

```{r}
#' Compute merge probabilities
#'
#' Computes merge probability vector as in Section 2. When tau is 0, puts equal
#' probability on all merges that have minimum dissimilarity.  Probabilities are
#' taken to be 0 on upper triangle and diagonal.
#'
#' @param d a symmetric k-by-k matrix of dissimilarities between clusters
#' @param tau amount of randomization (tau = 0 is without randomization)
get_merge_probs <- function(d, tau) {
  if (tau == 0)
    mat <- d == min(d[lower.tri(d, diag=FALSE)])
  else {
   # mat <- exp(-(d - max(d)) / tau)
    d_lower <- d[lower.tri(d, diag = FALSE)]
    tau_eff <- tau* mean(d_lower)
   mat <- exp(-d / tau_eff)
  }
  mat[upper.tri(mat, diag=TRUE)] <- 0
  return(mat / sum(mat))
}
```


Here's a small example to check that this is working as desired:

```{r}
d <- matrix(c(
  0, 1, 1, 2,
  1, 0, 3, 5,
  1, 3, 0, 6,
  2, 5, 6, 0
), nrow = 4, byrow = TRUE)
get_merge_probs(d, tau = 0)
round(get_merge_probs(d, tau = 1), 3)
```

Now we are ready to define the main function.

```{r}
#' Randomized Hierarchical Clustering
#' 
#' Performs hierarchical clustering with randomization and outputs an `hclust` 
#' object.
#' @param d a symmetric n-by-n matrix of dissimilarities
#' @param linkage_func a function taking two index sets in {1,...,n}
#' @param tau amount of randomization (tau = 0 is without randomization)
#' @export 
hclust_randomized <- function(d, linkage_func = linkage_complete, tau = 0.1) {
  if ("dist" %in% class(d)) d <- as.matrix(d)
  n <- nrow(d)
  clusts <- as.list(seq(n))
  clust_names <- -seq(n) # naming to match "merge" output of hclust
  d_clusts <- d
  merge <- matrix(0, n - 1, 2)
  height <- numeric(n - 1)
  for (i in seq(n - 1)) {
    # calculate probabilities and choose random pair to merge
    probs <- get_merge_probs(d_clusts, tau)
    probs_vec <- as.vector(probs)
    merge_idx <- sample(seq(length(probs_vec)), size = 1, prob = probs_vec)
    merge_clusts <- arrayInd(merge_idx, .dim = dim(d_clusts))
    c1 <- merge_clusts[1]
    c2 <- merge_clusts[2]
    
    # record height and merge for hclust object
    height[i] <- d_clusts[c1, c2]
    merge[i, ] <- sort(c(clust_names[c1], clust_names[c2]))    
    if (merge[i, 1] < 0 && merge[i, 2] < 0) {
      merge[i, ] <- rev(merge[i, ])
    }
    if (i == n-1) break # done with all n-1 merges
    
    # perform merge
    clusts <- append(clusts[-c(c1, c2)], list(c(clusts[[c1]], clusts[[c2]])))
    clust_names <- c(clust_names[-c(c1, c2)], i)
    d_clusts <- d_clusts[-c(c1, c2), -c(c1, c2), drop = FALSE]
    d_new <- numeric(length(clusts) - 1)
    for (j in seq(length(clusts) - 1)) {
      d_new[j] <- linkage_func(clusts[[length(clusts)]], clusts[[j]], d)
    }

    d_clusts <- rbind(d_clusts, d_new)
    d_clusts <- cbind(d_clusts, c(d_new, 0))
  }
  hc <- structure(list(
    merge = merge,
    height = height,
    labels = as.character(1:n),
    method = paste0("randomized (tau = ", tau, ")"),
    call = match.call(),
    dist.method = "user",
    linkage_func = linkage_func,
    tau = tau
    ), class = c("hclust_randomized", "hclust"))
  hc$order <- stats::order.dendrogram(stats::as.dendrogram(hc))
  return(hc)
}
```

Let's apply this on our example data.  We would like to check that when $\tau=0$, we get the same deterministic result as regular `hclust`.  We are also curious to see that when $\tau>0$, we get something different.

```{r}
# complete linkage
ex1$hc_complete <- list(
  hc_rand2 = hclust_randomized(ex1$d, linkage_func = linkage_complete, tau = 0.5),
  hc_rand1 = hclust_randomized(ex1$d, linkage_func = linkage_complete, tau = 0.1),
  hc_det = hclust_randomized(ex1$d, linkage_func = linkage_complete, tau = 0),
  hc = hclust(ex1$d, method = "complete")
)

# single linkage
ex1$hc_single <- list(
  hc_rand2 = hclust_randomized(ex1$d, linkage_func = linkage_single, tau = 0.5),
  hc_rand1 = hclust_randomized(ex1$d, linkage_func = linkage_single, tau = 0.1),
  hc_det = hclust_randomized(ex1$d, linkage_func = linkage_single, tau = 0),
  hc = hclust(ex1$d, method = "single")
)

is_hc_match <- function(hc1, hc2) {
  all(
    hc1$merge == hc2$merge,
    abs(hc1$height - hc2$height) < 1e-9,
    hc1$order == hc2$order
  )
}
is_hc_match(ex1$hc_complete$hc_det, ex1$hc_complete$hc)
is_hc_match(ex1$hc_single$hc_det, ex1$hc_single$hc)
```

We've verified above that for both single and complete linkage clustering, our function with $\tau=0$ gives the identical dendrogram as `hclust()`.

Let's look at how the randomized version clusters differently.

First, when $\tau=0.1$:

```{r}
plot_clustered_data(ex1$x,
                    clust1 = cutree(ex1$hc_complete$hc_det, k = 2),
                    clust1_label = ex1$hc_complete$hc_det$method,
                    cutree(ex1$hc_complete$hc_rand1, k = 2),
                    clust2_label = ex1$hc_complete$hc_rand1$method)
```

We see some borderline blue points are now circles rather than triangles.

Next, let's increase the randomization level to $\tau=0.5$:

```{r}
plot_clustered_data(ex1$x,
                    clust1 = cutree(ex1$hc_complete$hc_det, k = 2),
                    clust1_label = ex1$hc_complete$hc_det$method,
                    cutree(ex1$hc_complete$hc_rand2, k = 2),
                    clust2_label = ex1$hc_complete$hc_rand2$method)
```

We see even a greater discrepancy between the two clusterings.

