# Inference

We imagine a model in which in principle, each observation is distributed

$$
X_i\sim N_p(\mu_i,\sigma^2 I_p),
$$
and we would like to test whether all the points in the union of two sets are equal:

$$
H_0:\mu_i=\mu_{i'} \text{ for all }i,i'\in C_1\cup C_2.
$$

When $C_1$ and $C_2$ are predetermined sets, a ratio of the between to within variability follows an F-distribution.  In our case, we would like to compute p-values along each step of the randomized hierarchical clustering algorithm presented in the previous section.  This means that the sets $C_1$ and $C_2$ are themselves functions of the data, invalidating the classical F-test.  We will still use the same test statistic, but then compute the p-value differently.

## Computing statistics

Our test statistic is defined as the ratio of the between to within cluster sum of squared errors:

$$
R=(|C_1|+|C_2|-2)\cdot\frac{\|b\|^2}{\|w\|^2}
$$
where $b,w\in\mathbb R^{n\times p}$ have row $i$ given by

$$
b_i=\frac{\bar X_{C_1}-\bar X_{C_2}}{1/|C_1|+1/|C_2|}\cdot\begin{cases}1/|C_1|\text{ for }i\in C_1\\
-1/|C_2|\text{ for }i\in C_2\\0\text{ otherwise}\end{cases}
$$
and
$$
w_i=\begin{cases}X_i-\bar X_{C_1}\text{ for }i\in C_1\\
X_i-\bar X_{C_2}\text{ for }i\in C_2\\
0_p\text{ otherwise}
\end{cases}.
$$
We also note that

$$
\|b\|^2=\frac{\|\bar X_{C_1}-\bar X_{C_2}\|^2}{(1/|C_1|+1/|C_2|)^2}\cdot\left(|C_1|(1/|C_1|)^2+|C_2|(1/|C_2|^2)\right)=\frac{\|\bar X_{C_1}-\bar X_{C_2}\|^2}{1/|C_1|+1/|C_2|}
$$

In addition to computing the observed test statistic, we will also compute several "auxiliary statistics," whose observed values also depend on the matrices $b$ and $w$.

In particular, we will later need the matrix-valued function

$$
X(r';\mathcal A_o^{(t)})=\sqrt{\|b\|^2+\|w\|^2}\left(\frac{b}{\|b\|}\cdot g(r')+ \frac{w}{\|w\|}\sqrt{1-g(r')^2}\right) + X - b - w,
$$

$$
X(r';\mathcal A_o^{(t)})=\sqrt{1+1/a} \cdot g(r')\cdot b+\sqrt{1+a}\cdot\sqrt{1-g(r')^2}w+(X - b - w),
$$

where $g(r')=\sqrt{r'/(|C_1|+|C_2|-2+r')}$ and $a=\|b\|^2/\|w\|^2$.

Thus, our function will return $a, b$, and $w$ in addition to the test statistic.

Note: We will throw an error if someone attempts to test a merge of two singleton clusters.  This is excluded since with $\sigma$ unknown, this does not seem reasonable.

```{r}
#' Compute the test and information for auxiliary statistics
#' 
#' This computes the test statistic, which is the ratio of the squared Frobenius
#' norms of two matrices `b` and `w` times a scaling factor.  This test 
#' statistic is F-distributed when `clust1` and `clust2` are predetermined (i.e.
#' not learned on data being used for test).  But in our setting, the clusters 
#' are learned on the same data.  To properly account for this double dipping, 
#' we will also need the values of some additional statistics, which depend on 
#' `b` and `w`.
#' 
#' @param x n-by-p data matrix
#' @param clust1 a vector giving a subset of `1:n`
#' @param clust2 a vector giving a subset of `1:n`
compute_observed_statistics <- function(x, clust1, clust2) {
  if ("numeric" %in% class(x)) x <- matrix(x, ncol = 1)
  n1 <- length(clust1)
  n2 <- length(clust2)
  if (n1 == 1 & n2 == 1)
    stop("At least one of the tested clusters must be non-singleton")
  scaling <- 1/n1 + 1/n2
  p <- ncol(x)
  xx1 <- scale(x[clust1, , drop = FALSE], center = TRUE, scale = FALSE)
  xx2 <- scale(x[clust2, , drop = FALSE], center = TRUE, scale = FALSE)
  xbar_diff <- attr(xx1, "scaled:center") - attr(xx2, "scaled:center")

  # between <- sum(xbar_diff^2)
  # within <- sum(xx1^2) + sum(xx2^2)
  
  # compute w and b as sparse matrices
  w <- b <- Matrix::Matrix(0, nrow = nrow(x), ncol = p)
  w[clust1, ] <- xx1
  w[clust2, ] <- xx2

  b[clust1, ] <- 1/n1 * matrix(xbar_diff, nrow = n1, ncol = p, byrow = TRUE)
  b[clust2, ] <- -1/n2 * matrix(xbar_diff, nrow = n2, ncol = p, byrow = TRUE)
  b <- b / scaling # scaling = || nu ||^2
  
  a <- sum(b^2) / sum(w^2)
  ratio <- (n1 + n2 - 2) * a
  list(ratio = ratio, b = b, w = w, a = a)
}
```

```{r}
usethis::use_package("Matrix")
```


Let's test that this function works as expected.

```{r}
testthat::test_that("compute_observed_statistics() works", {
  # imagine a one-dim example with three points:
  x <- c(-1, 1, 6, 10)
  clust1 <- 1:2
  clust2 <- 3
  # working directly from the formula in the paper:
  scaling = 1/2 + 1/1
  bcss <- 2 * 1 / 3 * (0-6)^2
  wcss <- 1 + 1 + 0
  out <- compute_observed_statistics(x, clust1, clust2)
  testthat::expect_equal(out$ratio, (3 - 2) * bcss / wcss)
  testthat::expect_equal(out$a, sum(out$b^2) / sum(out$w^2))
  
  # and a two-dimensional example with four points:
  x <- matrix(c(-1, 1, 6, 6, 10,
                 0, 0, 2, 2, 10),
              ncol = 2)
  clust1 <- 1:2
  clust2 <- 3:4
  scaling = 1/2 + 1/2
  # working directly from the formula in the paper:
  bcss <- 2 * 2 / 4 * sum(c(6, 2)^2)
  wcss <- 1 + 1 + 0 + 0
  out <- compute_observed_statistics(x, clust1, clust2)
  testthat::expect_equal(out$ratio, (4 - 2) * bcss / wcss)
  testthat::expect_equal(out$a, sum(out$b^2) / sum(out$w^2))
})
```


## Conditionally valid p-value

According to Theorem 1, this involves computing the definite integral

$$
I(r)=\int_0^r \ell_{F}(r')\prod_{s=1}^tp^{(s)}\left(M_o^{(s)};X(r';\mathcal A_o^{(t)})\right)dr',
$$

where $\ell_{F}$ is the density of a $F_{p,(|C_1|+|C_2|-2)p}$ and $p^{(s)}$ is the merge probability defined in [this section](#randomized-hclust) evaluated at $X(r';\mathcal A_o^{(t)})$, which in the notation of the previous section is

$$
X(r';\mathcal A_o^{(t)})=\sqrt{1+1/a} \cdot g(r')\cdot b+\sqrt{1+a}\cdot\sqrt{1-g(r')^2}w+(X - b - w),
$$

To evaluate the integrand, we need to compute the chain of probabilities, from step 1 to $t$, of each merge that we made on the real data if $X(r';\mathcal A_o^{(t)})$ had been our data.  We will start by writing a function that takes the merge matrix from an `hclust` object and returns two things:

1. a list called `clusts` with $n-1$ elements, where the $i$-th element of the list represents the clusters that were present before the $i$-th merge occurs.

2. a matrix called `merges` whose $i$-th row gives the indices of the two clusters in `clusts[[i]]` that get merged at step $i$.


```{r}
#' Get sequence of clusters and merges from an `hclust` object
#' 
#' @param hc an object of class `hclust`
#' @return `clusts` is a list, where `clusts[[i]]` is the clustering before
#' merge i occurs.  The first cluster in `clusts[[i]]` is the one that was newly
#' created. `merges` is a matrix whose i-th row gives the indices of the
#' two clusters in `clusts[[i]]` being merged  
#' @export
get_hclust_sequence <- function(hc) {
  n <- nrow(hc$merge) + 1
  clusts <- list()
  clusts[[1]] <- as.list(1:n) # initially each leaf is in own cluster
  merges <- matrix(NA, n - 1, 2)
  cl <- list()
  for (i in seq(n - 1)) {
    for (j in 1:2) {
      if (hc$merge[i, j] < 0)
        cl[[j]] <- -hc$merge[i, j]
      else
        cl[[j]] <- clusts[[hc$merge[i, j] + 1]][[1]] # clust created at merge i
    }
    merges[i, ] <- match(cl, clusts[[i]])
    if (i == n - 1) break # no need to form a single cluster clustering
    unaltered_clusters <- clusts[[i]][-merges[i, ]]
    new_cluster <- sort(unlist(cl))
    clusts[[i+1]] <- c(list(new_cluster), unaltered_clusters)
  }
  list(clusts = clusts, merges = merges)
}
```

```{r}
testthat::test_that("get_hclust_sequence() works", {
  x <- c(1, 2,     10, 14,              40)
  n <- length(x)
  hc <- hclust(dist(x))
  out <- get_hclust_sequence(hc)
  
  # check that the clustering at each step is a partition of 1:n:
  for (i in seq(n-1)) testthat::expect_equal(sort(unlist(out$clusts[[i]])), 1:n)

  # check that the newly formed clusters are as expected:
  testthat::expect_equal(out$clusts[[2]][[1]], 1:2)
  testthat::expect_equal(out$clusts[[3]][[1]], c(3, 4))
  testthat::expect_equal(out$clusts[[4]][[1]], 1:4)

  # check that the clusters merged correctly form the new clusters:
  for (i in seq(n - 2)) {
    clust1 <- out$clusts[[i]][[out$merges[i, 1]]]
    clust2 <- out$clusts[[i]][[out$merges[i, 2]]]
    testthat::expect_equal(out$clusts[[i + 1]][[1]], sort(c(clust1, clust2)))
  }
})
```

### Forming integrand

We are now ready to write a function that will return the $t$ merge probabilities that get multiplied together in the integrand.

```{r}
#' Compute selection probabilities on a new data set
#' 
#' These probabilites are used in the integrand of the p-value calculation.
#' 
#' @param x_new observed data matrix
#' @param hc_sequence output of `get_hclust_sequence()` applied to output of
#' `hclust_randomized()`
#' @param dist_func the function that was used to compute the original n by n matrix
#' @param linkage_func the linkage used to measure the dissimilarity
#' between pairs of clusters
#' @param tau
compute_selected_probs <- function(x_new, hc_sequence, dist_func, linkage_func, tau) {
  num_steps <- length(hc_sequence$clusts) # different from nrow(x_new) - 1 if truncated in integrand
  probs_selected <- rep(NA, num_steps)
  
  # this part is much like `hclust_randomized()` in how it updates d_clusts
  # and computes probabilities, but it is different in that it goes with the
  # predetermined sequence of merges
  d <- as.matrix(dist_func(x_new))
  d_clusts <- d
  for (i in seq(num_steps)) {
    # calculate probabilities
    probs <- get_merge_probs(d_clusts, tau)
   
    # get the chosen two clusters from the current clustering
    c1 <- hc_sequence$merges[i, 1]
    c2 <- hc_sequence$merges[i, 2]
    # record probability for chosen pair
    probs_selected[i] <- probs[c1, c2] + probs[c2, c1] # matrix is lower tri
    if (i == num_steps) break

    # update dissimilarities
    d_clusts <- d_clusts[-c(c1, c2), -c(c1, c2), drop = FALSE]
    num_clusts <- length(hc_sequence$clusts[[i]])
    d_new <- c()
    for (j in seq(num_clusts)[-c(c1, c2)]) {
      d_new <- c(d_new, linkage_func(
        hc_sequence$clusts[[i + 1]][[1]], # newly created cluster
        hc_sequence$clusts[[i]][[j]], # remaining clusters
        d
      ))
    }

    # recall that new cluster will be first, i.e. clusts[[i+1]][[1]]:
    d_clusts <- rbind(d_new, d_clusts)
    d_clusts <- cbind(c(0, d_new), d_clusts)
  }
  
  return(probs_selected)
}
```


Let's make some tests based on the previous example:

```{r}
testthat::test_that("get_hclust_sequence() works", {
  x <- matrix(c(1, 2,     10, 14,              40), ncol = 1)
  n <- length(x)
  hc <- hclust(dist(x))
  out <- get_hclust_sequence(hc)
  
  # when tau = 0 and original x is used all probs should be 1 when there are 
  # no tied distances:
  testthat::expect_equal(
    compute_selected_probs(x, out, dist, linkage_complete, tau = 0),
    rep(1, n-1))
  
  # when tau = 0 and there are two tied distances:
  x_new <- matrix(c(1, 2, 3,     14,              40), ncol = 1)
  testthat::expect_equal(
    compute_selected_probs(x_new, out, dist, linkage_complete, tau = 0),
    c(0.5, 0, 1, 1))

  # when tau = 0 and there are three tied distances:
  x_new <- matrix(c(1, 2, 3, 4,              40), ncol = 1)
  testthat::expect_equal(
    compute_selected_probs(x_new, out, dist, linkage_complete, tau = 0),
    c(1 / 3, # merging 1-2 is as good as 2-3 or 3-4
      1, # merging 3-4: better than {1,2}-3 and {1,2}-4 for complete linkage
      1, # merging {1,2}-{3,4} is best
      1))
})
```

Recall that the integrand is given by 

$$
\ell_{F}(r')\prod_{s=1}^tp^{(s)}\left(M_o^{(s)};X(r';\mathcal A_o^{(t)})\right).
$$

It remains to multiply the $F_{p,(|C_1|+|C_2|-2)p}$ density by the $t$ probabilities.


```{r}
#' Integrand in p-value calculation
#' 
#' Creates a perturbed new data matrix based on the observed data and a 
#' perturbation of size r. Then `hc$merge` to reconstruct the possible merges at
#' each step s from 1 to t. Then, computes the probability of this sequence of 
#' merges being chosen for on this perturbed data set.
#' 
#' @param r scalar value at which we evaluate the integrand
#' @param t which step p-value is being calculated at
#' @param x observed data matrix
#' @param hc_sequence output of `get_hclust_sequence()` applied to output of
#' `hclust_randomized()`
#' @param obs_statistics output of `compute_observed_statistics()`
#' @param dist_func the function that was used to compute the original n by n matrix
#' @param linkage_func the linkage used to measure the dissimilarity
#' between pairs of clusters
#' @param tau
integrand <- function(r, t, x, hc_sequence, obs_statistics, dist_func, 
                      linkage_func, tau) {
  p <- ncol(x)
  # truncate sequence to step t:
  hc_sequence$clusts <- hc_sequence$clusts[1:t]
  hc_sequence$merges <- hc_sequence$merges[1:t, , drop = FALSE]
  # get the cluster sizes for the two clusters being tested:
  n1 <- length(hc_sequence$clusts[[t]][[hc_sequence$merges[t, 1]]])
  n2 <- length(hc_sequence$clusts[[t]][[hc_sequence$merges[t, 2]]])
  
  scaling = 1/n1+1/n2
  fr <- sqrt(r / (n1 + n2 - 2 + r))
  x_new <- with(obs_statistics, # all variables other than fr come from list
                sqrt(1+1/a) * fr * b + sqrt((1+a) * (1-fr^2)) * w + (x - b - w))

  probs <- compute_selected_probs(x_new, hc_sequence, dist_func, linkage_func, tau)
  return(df(r, df1 = p, df2 = p * (n1 + n2 - 2)) * prod(probs))
}
```




Let's make a plot of the integrand:

```{r}
x <- ex1$x
x <- matrix(c(1, 2,     10, 14,              40), ncol = 1)
set.seed(123)
hc <- hclust_randomized(dist(x), tau = 0.2)
hc_sequence <- get_hclust_sequence(hc)
tt <- 3 # last merge, when only two clusters remain
cl <- hc_sequence$clusts[[tt]] 
obs_stat <- compute_observed_statistics(x, cl[[1]], cl[[2]])

f <- function(r) integrand(r=r, t=tt, x=x, hc_sequence=hc_sequence,
                           obs_statistics=obs_stat,
                           dist_func = dist,
                           linkage_func = hc$linkage_func, tau = hc$tau)

rr <- seq(0, 10, length=40)
ff <- sapply(rr, f)
plot(rr,ff)
plot(rr,ff, log = "y")
```

We observe that it is monotonically increasing.

### Computing p-value

We are now ready to compute the integral $I(r)$, given at the top of the section.  By Corollary 1 of the paper, our selective p-value is given by 

$$
1 - \frac{I(r)}{I(\infty)}.
$$
```{r}
#' Compute selective p-value
#' 
#' Arguments `hc_sequence` and `obs_statistics` if `NULL` are computed from `hc`
#' and `t`.
#' 
#' @param t which step p-value is being calculated at
#' @param x observed data matrix
#' @param hc output of `hclust_randomized()`
#' @param hc_sequence output of `get_hclust_sequence()` applied to output of
#' `hclust_randomized()`
#' @param obs_statistics output of `compute_observed_statistics()`
#' @param dist_func the function that was used to compute the original n by n matrix
#' 
#' @export
compute_pvalue <- function(t, x, hc, hc_sequence = NULL, obs_statistics = NULL, dist_func) {
  stopifnot("hclust_randomized" %in% class(hc))
  
  if (is.null(hc_sequence)) hc_sequence <- get_hclust_sequence(hc)
  if (is.null(obs_statistics)) {
    # at step t, these are the two clusters that get merged:
    clust1 <- hc_sequence$clusts[[t]][[hc_sequence$merges[t, 1]]]
    clust2 <- hc_sequence$clusts[[t]][[hc_sequence$merges[t, 2]]]
    obs_statistics <- compute_observed_statistics(x, clust1, clust2)
  }

  integrand_func <- Vectorize(function(r) {
    integrand(r=r,
              t=t,
              x=x,
              hc_sequence=hc_sequence,
              obs_statistics=obs_statistics,
              dist_func = dist_func,
              linkage_func = hc$linkage_func,
              tau = hc$tau)
  })
  # rather than 1 - I(r)/I(infty), let's try 1 / [1 + I(r) / I(r to inf)]
  numerator <- stats::integrate(integrand_func, 0, obs_statistics$ratio)
  extra <- stats::integrate(integrand_func, obs_statistics$ratio, Inf)
  return(1 - 1 / (1 + extra$value / numerator$value))
}
```

Let's try this out on our example.

```{r}
x <- ex1$x
set.seed(123)
hc <- hclust_randomized(dist(x), tau = 0.2)
tt <- 50
pval <- compute_pvalue(tt, x, hc, dist_func = dist)
```

Let's see if the p-values have a uniform distribution under the global null:

```{r}
set.seed(123)
n <- 60
p <- 2
tt <- 59 # test when there are only two clusters
tau <- 0.05
nsim <- 100
pvals <- rep(NA, nsim)
for (i in seq(nsim)) {
  x <- matrix(rnorm(n * p), n, p)
  hc <- hclust_randomized(dist(x), tau = tau)
  pvals[i] <- compute_pvalue(tt, x, hc, dist_func = dist)
}
```

```{r}
pvals
```

```{r}
qqplot(
  (ppoints(length(pvals))),
  (sort(pvals)),
  xlab = "Expected",
  ylab = "Observed"
)
abline(0, 1, col = "red")
```

Looks good.

#### An alternative grid-based approach

Let's try out the adaptive grid-based summation approach that was initially used in the Python implementation:

Suppose we have a pdf $f$ that we want to integrate, but we don't know where its support is concentrated and its costly to evaluate.  Our approach is the following:

- compute $f$ along a wide coarse grid.
- interpolate $f$ along a finer version of this wide grid
- evaluate cdf of interpolated version of $f$ to get 0.005 and 0.995 quantiles
- get a narrower grid based on this (going 3x wider than the quantiles)
- on the narrower grid, evaluate on a coarse grid
- now, interpolate using this narrow coarse grid on a narrow fine grid to get p-value

```{r}
#' Adaptive grid integration for pdf evaluation
#'
#' Performs two-pass adaptive grid refinement to efficiently evaluate a 
#' probability density function. First identifies the main part of support on a 
#' wide grid, then refines evaluation on a narrower grid focused on the 
#' high-density region.
#'
#' @param f A function that takes a single numeric input and returns the 
#'   (possibly unnormalized) probability density at that point.
#' @param observed_target Numeric. The observed value of the test statistic 
#'   for which to compute a p-value.
#' @param wide_grid_min Numeric. The minimum value of the initial wide grid.
#' @param wide_grid_max Numeric. The maximum value of the initial wide grid.
#' @param num_coarse Integer. The number of points to use in the coarse grids 
#'   for both passes.
#' @param num_fine Integer. The number of points to use in the fine grid for 
#'   the first pass. The second pass uses \code{num_fine * 2} points.
#' @param qlow Numeric. The lower quantile for defining the narrower interval 
#'   (default: 0.005).
#' @param qhigh Numeric. The upper quantile for defining the narrower interval 
#'   (default: 0.995).
#' @param buffer Numeric. Factor by which to expand the narrower interval width 
#'   on each side. The final grid spans from `q_low - buffer * width` to 
#'   `q_high + buffer * width`.
adaptive_integrate <- function(f, observed_target, wide_grid_min, wide_grid_max, 
                               num_coarse, num_fine, 
                               qlow = 0.005, qhigh = 0.995, buffer = 3) { 
  
  # Pass 1: Wide grid
  wide_fine_grid <- seq(wide_grid_min, wide_grid_max, length.out = num_fine)
  
  # Evaluate f on fine grid
  f_fine <- sapply(wide_fine_grid, f)
  f_fine <- pmax(0, f_fine)
  
  # Normalize pdf using trapezoidal rule
  dx <- diff(wide_fine_grid)
  total_mass <- sum((f_fine[-1] + f_fine[-length(f_fine)]) / 2 * dx)
  f_fine_norm <- f_fine / total_mass
  
  # Compute CDF using trapezoidal rule
  cdf <- c(0, cumsum((f_fine_norm[-1] + f_fine_norm[-length(f_fine_norm)]) / 2 * dx))
  
  # Remove duplicates for interpolation
  unique_idx <- !duplicated(cdf)
  cdf_unique <- cdf[unique_idx]
  grid_unique <- wide_fine_grid[unique_idx]
  
  # Find narrower interval
  low <- approx(cdf_unique, grid_unique, xout = qlow, rule = 2)$y
  high <- approx(cdf_unique, grid_unique, xout = qhigh, rule = 2)$y
  width <- high - low
  low <- max(wide_grid_min, low - buffer * width)
  high <- min(wide_grid_max, high + buffer * width)
  
  cat("Narrow grid bounds:", low, high, "\n")
  
  # Pass 2: Narrow coarse grid
  narrow_coarse_grid <- seq(low, high, length.out = num_coarse)
  
  # Evaluate f on narrow coarse grid
  f_narrow_coarse <- sapply(narrow_coarse_grid, f)
  
  # Interpolate to narrow fine grid
  narrow_fine_grid <- seq(low, high, length.out = num_fine * 2)
  interp_func_narrow <- splinefun(narrow_coarse_grid, f_narrow_coarse, method = "natural")
  f_narrow_fine <- pmax(0, interp_func_narrow(narrow_fine_grid))
  
  # Compute p-value
  # Normalize the posterior
  posterior <- f_narrow_fine
  
  #posterior <- posterior/max(posterior)
  posterior_sum <- sum(posterior)
  
  if (posterior_sum == 0 || is.nan(posterior_sum)) {
    # Fallback: use uniform distribution to avoid NaN
    posterior <- rep(1 / length(posterior), length(posterior))
    posterior_sum <- sum(posterior)
  } else {
    posterior <- posterior / posterior_sum
  }
  
  # Calculate p-value: P(statistic >= observed_target)
  p_value <- sum(posterior[narrow_fine_grid >= observed_target])

  # Return results
  return(list(
    p_value = p_value,
    grid = narrow_fine_grid,
    pdf = f_narrow_fine,
    narrow_bounds = c(low, high)
  ))
}
```

```{r}
f <- function(x) dnorm(x, mean = 25, sd = 2)

prob <- 0.95

result <- adaptive_integrate(
  observed_target = qnorm(prob, mean = 25, sd = 2),
  f = f,
  wide_grid_min = 0,
  wide_grid_max = 700,
  num_coarse = 20,
  num_fine = 1000,
  buffer = 0.5 # needed a narrower buffer for this to work
)

# Plot the result
plot(result$grid, result$pdf, type = "l", 
     xlab = "x", ylab = "pdf",
     main = "Adaptively refined pdf")

result$p_value - (1-prob) # pretty close!
```



```{r}
#' Compute selective p-value
compute_pvalue_grid <- function(t, x, hc, hc_sequence = NULL, obs_statistics = NULL, dist_func) {
  stopifnot("hclust_randomized" %in% class(hc))
  
  if (is.null(hc_sequence)) hc_sequence <- get_hclust_sequence(hc)
  if (is.null(obs_statistics)) {
    # at step t, these are the two clusters that get merged:
    clust1 <- hc_sequence$clusts[[t]][[hc_sequence$merges[t, 1]]]
    clust2 <- hc_sequence$clusts[[t]][[hc_sequence$merges[t, 2]]]
    obs_statistics <- compute_observed_statistics(x, clust1, clust2)
  }
  integrand_func <- Vectorize(function(r) {
    integrand(r=r,
              t=t,
              x=x,
              hc_sequence=hc_sequence,
              obs_statistics=obs_statistics,
              dist_func = dist,
              linkage_func = hc$linkage_func,
              tau = hc$tau)
  })
  
  results <- adaptive_integrate(
    observed_target = obs_statistics$ratio,
    f = integrand_func,
    wide_grid_min = 0.00001,
    wide_grid_max = 180,
    num_coarse = 200, # making this larger improves approximation, but at cost to time
    num_fine = 2000,
    buffer = 3
    )
  
  return(results$p_value)
}
```


Let's see if the p-values have a uniform distribution under the global null:

```{r}
n <- 30
p <- 2
tt <- 29 # test when there are only two clusters
tau <- 0.05
nsim <- 5

# run compute_pvalue() 
set.seed(123)
pvals <- rep(NA, nsim)

time1 <- system.time({
for (i in seq(nsim)) {
  x <- matrix(rnorm(n * p), n, p)
  hc <- hclust_randomized(dist(x), tau = tau)
  pvals[i] <- compute_pvalue(tt, x, hc, dist_func = dist)
}
})

# run compute_pvalue_grid() on identical data sets 
set.seed(123)
pvals2 <- rep(NA, nsim)
time2 <- system.time({
for (i in seq(nsim)) {
  x <- matrix(rnorm(n * p), n, p)
  hc <- hclust_randomized(dist(x), tau = tau)
  pvals2[i] <- compute_pvalue_grid(tt, x, hc, dist_func = dist)
}
})

pvals2
```


```{r}
qqplot(
  (ppoints(length(pvals2))),
  (sort(pvals2)),
  xlab = "Expected",
  ylab = "Observed",
  main = "Grid-based"
)
abline(0, 1, col = "red")
```

In principle, this should be identical to the p-values we got with the other approach.  Let's check how close they are:

```{r}
range(pvals-pvals2)
time1
time2
```

This shows that we can get nearly the identical p-values, but that the grid-based approach may be substantially slower.  Note: We can speed up the grid-based approach by taking `num_coarse` to be smaller; however, then the approximation suffers.

For this reason, we will (for now) export `compute_pvalue()` and not the grid-based approach.
